---
title: "Projet 1"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
```{r include=FALSE}
library(tidyverse);
library(scales);
library(lubridate);
library(ggmap);
library(ggplot2);
library(sp);
```


## Exploring the dataset
We were given the dataset *2016collision.csv*. If we inspect the file, we can see that we have 14023 observations of 13 variables. Some of those variable are categorical, but other are not, so we need to be careful when we import the dataset so that R doesn't convert numerical variable into factor.

```{r}
df <- read.csv(file = '2016collisionsfinal.csv', header = T, stringsAsFactors = c(0,0,0,0,0,0,1,1,1,1,1,1,1))
head(df)
```
And the name of the 13 variables are:
```{r}
colnames(df)
```

### The variables
Here is a short interpretation of each variable:

  - **Record**: a unique number, ranging from 1 to 14023, given to each observation.
  - **Location**: the location, in the form of an address.
  - **X,Y**: the longitude and latitude of the collision based on the Universal Transverse Mercator (UTM) coordinate system.
  - **Date**: the date of the collision (in the format '%m/%d%y')
  - **Time**: the time of the collision (in the format '%H:%M')
  - **Environment**: a categorical variable that specifies the type of weather at the moment of the accident. It has levels:
```{r}
levels(df$Environment)  
``` 
Most of the levels are self-explanatory.
  
  - **Road_Surface**: a categorical variable that specify the condition of the road at the moment of the collision. It as levels:
```{r}
levels(df$Road_Surface)
```

  - **Traffic_Control**: a categorical variable that specifies the type of traffic control in place at the time of the accident. When we take a look at the levels of this variable, we can see that one is labeled *""* (i.e. a null string). Further inspection maybe required to determine if this is due to a missing value or if this is an actual level.
```{r}
levels(df$Traffic_Control)
```

  - **Collision_Location**: a categorical variable that specifies where the collision happened, but in terms of road types. We again got the label *""*, however we also have two labels for other road types: *"98 - Other"* and *"99 - Other"*. This is probably a data entry error, because all the other variables use the label *"99 - Other"*.
```{r}
levels(df$Collision_Location)
```

 - **Light**: a categorical variable that specify the amount of light (natural of artificial) at the moment of the collision. We again got the *""* label.
```{r}
levels(df$Light)
```

Since there are only two observations in the "" label, we may assume that these are missing values or a data entry errors.

- **Collision_Classification**: a categorical variable with 3 levels:
    - *01 - Fatal injury*: Implies that one or more person has decease due to the collision.
    - *02 - Non-fatal injury*: Implies that one or more person got injured as a result of the collision, but no one died.
    - *03 - P.D. only*: Implies that only some property damage was cause by the collision.
    
  - **Impact_type**: a categorical variable that specifies the type of impact for the collision. It has levels:
```{r}
levels(df$Impact_type)
```
### Simple plot of the categorical variables 

#### Environement
```{r}
p <- ggplot(data = df, aes(x = df$Environment)) + geom_bar()
show(p)
```

The labels are hard to read, rotating the plot might be a good idea.
```{r}
p + coord_flip() + xlab("Environment") 
```

We can see that the most recorded collision has occurs in a clear environment. 

### Road Surface

We can do the same with the road surface

```{r}
p <- ggplot(data = df, aes(x=df$Road_Surface)) + geom_bar() + coord_flip() + xlab("Road Surface")
show(p)
```
Accordingly to the result we got with the Environment variable, most collision happen on a dry surface.

### Trafic Control
```{r}
p <- ggplot(df, aes(x = df$Traffic_Control)) + geom_bar() + coord_flip() + xlab("Traffic control")
show(p)
```
In this plot, we can see that most collision happen when there were no traffic control, but a lot of accident also happen at a traffic light or at a stop.

### Collision Location
```{r}
p <- ggplot(df, aes(x = df$Collision_Location)) + geom_bar() + coord_flip() + xlab('Collision location')
show(p)
```

### Light
```{r}
p <- ggplot(df, aes(x = df$Collision_Location)) + geom_bar() + coord_flip() + xlab('Collision location')
show(p)
```

### Collision classification
```{r}
p <- ggplot(df, aes(x = df$Collision_Classification)) + geom_bar() + xlab("Collision classification")
show(p)
```

### Impact Type
```{r}
p <- ggplot(df, aes(df$Impact_type)) + geom_bar() + coord_flip() + xlab("Impact type")
show(p)
```

### Collision w.r.t the time of the day
Since we have access to the time of the collision, one could ask at what time does most collision happen? 

To do this, we need to convert the Time vector of string into something more useful, such as a vector of datatime object. 
```{r}
df$Time <- as.POSIXct(df$Time, format = '%H:%M')
plot_lim <- as.POSIXct(c('00:00','23:59'), format = '%H:%M')
```

```{r}
p <- ggplot(data = df, aes(x = Time)) + geom_histogram(binwidth = 1000)
show(p)
```

The first thing to say about this plot is that it is very unlikely that a number this large of collision happen exactly at 00:00. This could indicate many thing but most likely that some entries are missing or ill formatted.

```{r}
p + scale_x_datetime(limits =plot_lim, breaks=date_breaks("2 hours"), labels=date_format("%H:%M"))
```
While trying to fix the range and format of the x axis, for some reason, *geom_bar* automatically remove the missing value. But now, for some reason, the x axis limits goes from 5am to 4:59am the next day.

### Colision along the year
```{r}
df$Date<-as.Date(df$Date,format="%m/%d/%Y")
df$months<-month(df$Date)

x<-c(replicate(12,0))

for (i in 1:length(df$months)){
    for(j in 1:12)
        if(df$months[i]==j){
            x[j]=x[j]+1
        }
    }

y<-matrix(nrow=12,ncol=2)
for (i in 1:12){
    y[i,1]=i
    y[i,2]=x[i]
}

plot(y[1,],y[2,])


dataweekly<-week(df$Date)
 datayearly<-year(df$Date)
str(df)
summary(df)
hist(summary(df)[,5])
```
** In not sure why, but I can't run this code. I have made pretty much the same graph, but without doing database manipulations**
```{r}
p <- ggplot(df, aes(Date)) + geom_histogram(binwidth = 5)
show(p)
```



### Map Visualization

Since this dataset represents collisions in Ottawa that happened in 2016, we can plot the collisions on a map, using the coordinates (variables X and Y) from the dataset; this way, we can observe spatial details of the collisions data. 

Since the coordinates in the dataset were recorded according to the Modified Transverse Mercator coordinate system, we need to convert the coordinates to the latitude-longitude system since the Ottawa map is scaled according to the lat-long system.

First, we have to determine if there are any missing coordinates, because the converting function doesn't work if there are missing values :

```{r}

df$X <- as.numeric(df$X)
df$Y <- as.numeric(df$Y)

#finding records with missing values (when in numeric form)

for (i in 1:14023) {
  if (is.na(df$X[i])) {
    print(i)
  }
}
```

With a little bit more investigating, we see that the records with missing coordinates are actually entries in a different format.(ex: 356,762.29 instead of 356762.29) Therefore, we will manually input the correctly formatted values for these entries:

```{r}
#manually change the format of entries with different formats 

df[5150,3] <- "356762.29"
df[5150,4] <- "5015593.96"

df[5666,3] <- "365191.84"
df[5666,4] <- "5011253.07"

df[5942,3] <- "384542.73"
df[5942,4] <- "5034620.66"

df[5943,3] <- "384542.73"
df[5943,4] <- "5034620.66"

df[9238,3] <- "384661.88"
df[9238,4] <- "5034144.30"

df[12111,3] <- "356864.55"
df[12111,4] <- "5015648.07"

df[12290,3] <- "382197.79"
df[12290,4] <- "5033332.03"

df[13828,3] <- "356535.81"
df[13828,4] <- "5015509.05"

#change X and Y character to num again

df$X <- as.numeric(df$X)
df$Y <- as.numeric(df$Y)

```

Now, since we fixed the missing values, we can convert the coordinates to lat-long :

```{r}
#make variable of coords
coords <- cbind(Easting = df[,3], Northing = df[,4])

#make dataset without coords
data_less_points <- df[,-(3:4)]

#make data frame of spatial coords
utms <- SpatialPointsDataFrame(coords, data = data_less_points, proj4string = CRS("+init=epsg:2018 +datum=NAD83"))

#for conversion
latlong = "+init=epsg:4269"

#transforming mtms to lat-long
gps_coords <- spTransform(utms, CRS(latlong))

#making new data.frame with lat-long coords and all other variables
data_with_lat.long <- data.frame(gps_coords)
data_with_lat.long <- data_with_lat.long[,1:13]
names(data_with_lat.long) <-c("Record", "Location", "Date", "Time", "Environment", "Road_Surface", "Traffic_Control", "Collision_Location", "Light", "Collision_Classification", "Impact_Type", "Longitude", "Latitude")

```

Finally, the coordinates are converted to the lat-long system :

```{r}

head(data_with_lat.long$Longitude)

```

Next, we can get map tiles of Ottawa from Stamen Maps. We can change the size of the maps to focus on different areas of the city: 

(Map tiles by Stamen Design, under CC BY 3.0. Data by OpenStreetMap, under ODbL.)

```{r message=FALSE, warning=TRUE}

#getting tiles from Stamen and creating different maps

ottawa.map <- get_stamenmap(bbox = c(left = -76.5, bottom = 45.0, right = -75.0, top = 45.8), zoom = 11, maptype = c("terrain-lines"), messaging = FALSE)

ottawa.map2 <- get_stamenmap(bbox = c(left = -77.1384, bottom = 45.14344, right = -76.6394, top = 45.5506), zoom = 11, maptype = c("terrain-lines"), messaging = FALSE)

ottawa.map3 <- get_stamenmap(bbox = c(left = -76.2384, bottom = 45.14344, right = -75.2, top = 45.5506), zoom = 11, maptype = c("terrain-lines"), messaging = FALSE)

ottawa.map4 <- get_stamenmap(bbox = c(left = -76.3384, bottom = 45.0, right = -75.0, top = 45.5506), zoom = 11, maptype = c("terrain-lines"), messaging = FALSE)

ottawa.map.zoom <- get_stamenmap(bbox = c(left = -75.8, bottom = 45.3, right = -75.6, top = 45.45), zoom = 13, maptype = c("terrain-lines"), messaging = FALSE, color="color")

ottawa.map.zoomer <- get_stamenmap(bbox = c(left = -75.75, bottom = 45.38, right = -75.65, top = 45.44), zoom = 14, maptype = c("toner"), messaging = FALSE, color="color")

```

##Maps

Finally, we are ready to plot the collisions on the map.  

This is a broad view of the city, including the suburbs : 

```{r warning=FALSE}
#broad view of Ottawa
ggmap(ottawa.map)+ geom_point(data = data_with_lat.long, aes(x=Longitude, y=Latitude), size=0.5, col='blue') + scale_y_continuous(limits=c(45.0,45.55)) +
  scale_x_continuous(limits=c(-76.4,-75.0)) + labs(x="Longitude", y="Latitude") + ggtitle("Vehicle Collisions in Ottawa, in 2016")
```

We can see from the above plot that there are more collisions in the more populated areas of the city (downtown, Orléans, Barhaven, Kanata...) which is not really surprising. We may assume that, since there is a higher volume of vehicles on the road in more urban areas, there will be more collisions compared to more rural areas. It is also interesting to see the shape of the city, since the city limits are clearly distinguishable by the distribution of the collisions.

We can change the map by using a different scale, and get a closer look at the city:

```{r warning=FALSE}
#getting a better zoom

ggmap(ottawa.map3)+ geom_point(data = data_with_lat.long, aes(x=Longitude, y=Latitude), size=0.5, col="darkolivegreen") + scale_y_continuous(limits=c(45.15,45.55)) + scale_x_continuous(limits=c(-76.0,-75.4)) + labs(x="Longitude", y="Latitude")+ labs(x="Longitude", y="Latitude") + ggtitle("Vehicle Collisions in Ottawa, in 2016")
```


Again, we can clearly see that there are more collisions on the main roads, with the concentration of collisions increasing as the roads get closer to downtown Ottawa or to a suburb.

We can also observe the distribution of the collisions in Downtown Ottawa by using a map with a different scale :  

```{r warning=FALSE}
#view of DowntownOttawa

ggmap(ottawa.map.zoomer)+ geom_point(data = data_with_lat.long, aes(x=Longitude, y=Latitude), size=0.3, col = "deepskyblue") + scale_y_continuous(limits=c(45.38,45.44)) + scale_x_continuous(limits=c(-75.75,-75.65)) + ggtitle("2016 Collisions in Downtown Ottawa") + labs(x="Longitude", y="Latitude")

```

In this case, we can see that the roads leading into the downtown core, like Bank Street or Bronson Avenue, have a high frequency of collisions, presumably because these are typical roads for commuters and workers coming into town. Also, with this map, it is interesting to see that some intersections, like Elgin Street and Laurier Avenue West, or Slater Street and Lyon Street North, have a higher density of collisions points than other intersections, than say Laurier Avenue West and Bank Street. Possibly, some intersections may be dealing with a higher volume of vehicles than others, or maybe the configuration of the intersections may lead to it being more collision prone. 

Thus, these maps can help us identify possibly problematic areas with a many collisions points that would require further investigation.


###Bivariate Visualizations

Furthermore, other variables from the dataset plotted pairwise can give us a high-level understanding and overview of possible trends or patterns in the data. The next step in our data exploration will be to look at some bivariate plots.

###Time Variables

In the collisions dataset, we got variables for the time of the collision and the date of the collision. It may be interesting to see how these time variables interact with each other, and with other variables.

#####1) Hour and Month
First, the hour and the month of each collision could be of interest, therefore we extract this information from the *Time* and the *Date* variables :  

```{r}
library(ggpubr)
library(lubridate)
library(hexbin)
#library(plyr)
library(reshape2)

#extracting the month out of the date variable and making a new variable for month

df$Month <- as.factor(month(as.POSIXlt(df$Date, format="%m/%d/%Y")))
levels(df$Month) <- month.abb[1:12]

#extracting the hour out of the time variable and making a new variable for hour

df$Hour <- as.factor(format(as.POSIXct(df$Time,format="%H:%M"),"%H"))


```

Now, since we have the necessary information, we can see if there is any trends. The graphs below let us compare the total number of collisions by hour for every month of the year. 

```{r}
ggplot(df) + geom_bar(aes(Hour)) + facet_wrap(~Month, scales="free_y") + coord_flip() + theme(axis.text.y=element_text(size=rel(0.5)))+ggtitle("Number of Collisions per Hour for Every Month of the Year") + labs(x="Hour of the Day", y="Number of Collisions")
```

When we look at the above plots, there is a similar trend throughout the first months of the year (January, February, and March) and the last months of the year (September, October, November and December) : the distribution of the collisions is bimodal, meaning that there seems to be a higher number of collisions at two times in the day. This is expected and in line with the common knowledge of rush-hour traffic in the morning, and in the afternoon in a city like Ottawa.

On the other hand, for the typical "summer" months (April, May, June, July and August), the distribution of the collisions by hour seems to be more unimodal, with a peak during afternoon rush hour. There is still a peak of collisions during the morning rush-hours, but they are not as prominent in the summer months. This may be due to the fact that Ottawa residents typically take vacations during the summer months and maybe are on the roads later in the morning, or work hours may be less rigid during the summer, and workers go into work at later hours, leading to a more uniform number of vehicles on the road in the morning hours, leading to a more uniform number of collisions.

Also, if we specifically look at the peak collision hours in the afternoon/evening throughout the year, we see that generally, the peak collision hour for Jan-Apr is 16:00-17:00, while it is 15:00-16:00 for May-Sept, and again 16:17 for Oct-Dec. Similarly, if we look at the peak collision hours in the morning throughout the year, we generally see that during the winter months, the peak morning hour collision time is earlier in the morning (7:00-8:00), while in the summer months, the morning peak hour collision time is later in the morning (8:00-9:00).Again, these observations are not really surprising, if we assume that people may leave later for work and leave earlier from work during the summer months, leading to more vehicles on the road during these times.



```{r}
grp_month <- group_by(df,Month)
sum_month <- summarise(grp_month, mean = mean(Time), sd = sd(Time))
ggplot(sum_month,aes(Month,mean, group = 1)) + geom_point() + geom_line() + labs(title = "Average time of collision per month", y = "Mean")
```


```{r}
df$Week <- week(df$Date)
grp_week <- group_by(df,Week)
sum_week <- summarise(grp_week,mean = mean(Time))
ggplot(sum_week,aes(Week,mean, group = 1)) + geom_point() + geom_line() + labs(title = "Average time of collision per week", y = "Mean")
```

#####2) Month and Light Type

In the following plot, we will see the proportion of collisions that occurred during different lighting types, by month.

```{r}
#Light and Month

ggplot(df) + geom_bar(aes(Month, fill=Light),position="fill") + theme(axis.text.x=element_text(angle=0)) + ggtitle("Proportion of Accidents by Light type per Month") + labs(x="Month", y="Proportion of Collisions")
```

Not surprisingly, we see what we would expect to see, knowing how sunlight behaves throughout the year in Canada: a higher proportion of accidents happen during the daylight during the summer months, and in darkness during the winter months, since there are more daylight hours during the summer and darkness hours during the winter months.

#####3) Month and Environment Type/Month and Road_Surface

Next, we can plot the distribution of collisions by *Environment* and *Month*, as well as *Road Surface* and *Month* . Again, we expect to see results aligned with common sense. 

```{r}
#Month and Environment
ggplot(df) + geom_bar(aes(Month, fill=Environment),position="fill") + ggtitle("Proportion of Accidents by Environment Type per Month") + labs(x="Month", y="Proportion of Collisions")

#Month and Environment
ggplot(df) + geom_bar(aes(Month, fill=Road_Surface),position="fill") + ggtitle("Proportion of Accidents by Road Condition Type per Month") + labs(x="Month", y="Proportion of Collisions")
```

In fact, we see result that we expected. For instance, in Dec-Jan-Feb, between 20% and 40% of collisions happened when it was snowing. However, a bit unusually, there were more accidents in April when it was snowing than in March. Since there usually is less snow in April than in March, this is a interesting. Perhaps winter tires were no longer on the cars, which resulted in more collisions during snowy conditions? Also, fog/mist/smoke/dust seemed to be more present during collisions in November. 

The results for the road surface are equally predictable. In the summer months, 80-90% of collisions happened when the road was dry, while 15-20% of collisions in December and February happened when the road was icy. 

These observations are all in line with basic knowledge of Canadian driving conditions. Later on, we will plot the collisions on the map and identify them by road_surface and environment conditions, which can maybe give us a sense or help us target specific roads that may be problematic in terms of road conditions. 


###Other Variables

We will also try to observe the relationship between other variables. 

#####1) Impact type and Collision Classification

This graph separates the collisions by classification type (fatal injury, non-fatal injury and property damage only) and indicates the proportion of accidents by impact type:

```{r}
ggplot(df) + geom_bar(aes(Collision_Classification, fill=Impact_type),position="fill") + ggtitle("Proportion of Collision by Classification and Impact Type") + labs(x="Collision Classification", y="Proportion of Collisions")
```

From the graph above, we see that collisions that resulted in fatal injuries usually involved a SMV (single motor vehicle), a turning movement or an approaching vehicle. On the other hand, non-fatal and p.d only collisions were often rear end collisions. 

#####2) Impact type and Traffic_Control

```{r}
ggplot(df) + geom_bar(aes(Impact_type)) + 
  facet_wrap(~Traffic_Control, scales="free_y") + theme(axis.text.x=element_text(angle=90)) + ggtitle("Collisions by Impact Type per Traffic Control") + labs(x="Impact Type", y="Collisions")
```

From the above graphs, it is interesting to see that rear end collisions is the most frequent impact type at roundabouts, while angle collisions are the most common form of collisions at stop signs. We may want to reflect on why this is.

Also, we have to be careful in the interpretation of the pedestrian crossover graph. Since there was only one collision involving a pedestrian crossover, the fact that 100% of pedestrian crossover collisions were rear end collision does not mean anything, since it is only based on one observation.

#####3) Light and Collision Classification

In the next graph, we will try to determine links between the type of light at the time of the collision, and the collision classification:

```{r}
ggplot(df) + geom_bar(aes(Light)) + 
  facet_wrap(~Collision_Classification, scales="free_y") + 
  theme(axis.text.x=element_text(angle=90))+
  scale_fill_brewer(palette = "Set1", name = "Light") + ggtitle("Number of Accidents by Light Type per Collision Classification") + labs(x="Light Type", y="Number of Collisions")
```

Across the different collision classifications, the trend regarding the light is quite evident : most accidents in all classification types occur during the daylight, which makes sense since there are presumably more cars on the road at this time. Most of the collisions that don't happen during daylight happen when it is dark outside.


### Multivariate Visualizations - Maps

I need to add comments.

```{r}
ggmap(ottawa.map)+ geom_point(data = data_with_lat.long, aes(x=Longitude, y=Latitude, color=Light), size=0.5) + scale_y_continuous(limits=c(45.0,45.55)) +
  scale_x_continuous(limits=c(-76.4,-75.0)) + labs(x="Longitude", y="Latitude")+ ggtitle("2016 Collisions in Downtown Ottawa by Environment Type")

#zoomed onto downtown core with colours being different environments 
ggmap(ottawa.map.zoomer)+ geom_point(data = data_with_lat.long, aes(x=Longitude, y=Latitude, col=Environment), size=0.5) + scale_y_continuous(limits=c(45.38,45.44)) +
  scale_x_continuous(limits=c(-75.75,-75.65)) + ggtitle("2016 Collisions in Downtown Ottawa by Environment Type") + labs(x="Longitude", y="Latitude")

ggmap(ottawa.map.zoomer)+ geom_point(data = data_with_lat.long, aes(x=Longitude, y=Latitude, color=Traffic_Control), size=0.5) + scale_y_continuous(limits=c(45.38,45.44)) +
  scale_x_continuous(limits=c(-75.75,-75.65)) + ggtitle("2016 Collisions in Downtown Ottawa by Environment Type") + labs(x="Longitude", y="Latitude")


```

### Word Map

Since we have the name of each street/intersection of were each collision has happen, we can try to do some word cloud map to see which road came up more often. 
```{r}
library("tm")
library("wordcloud")
library("RColorBrewer")
```

```{r}
docs <- Corpus(VectorSource(df$Location))
docs <- tm_map(docs, tolower)
dtm <- TermDocumentMatrix(docs)
m <- as.matrix(dtm)
v <- sort(rowSums(m),decreasing=TRUE)
d <- data.frame(word = names(v),freq=v)
set.seed(1)
wordcloud(words = d$word, freq = d$freq, min.freq = 1,
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))
```
From this, we can see that the most used word is btwn, which doesn't gives much more insight other than how the collision are reported. But, we can also see that word like ave, blvd, highway are frequent, telling us that more collision were reported on those type of road.

We could try to remove some popular road type name, in order to get a word cloud containing only the road name.
```{r}
docs_road <- tm_map(docs, removeWords, c("ave", "btwn", "blvd", "pkwy", "highway", "st."))
dtm <- TermDocumentMatrix(docs_road)
m <- as.matrix(dtm)
v <- sort(rowSums(m),decreasing=TRUE)
d <- data.frame(word = names(v),freq=v)
set.seed(1)
wordcloud(words = d$word, freq = d$freq, min.freq = 1,
          max.words=200, random.order=FALSE, rot.per=0.35, 
          colors=brewer.pal(8, "Dark2"))
```

What is interesting here is that that road name that came up more are the same that have a high collision density on the previous map plot.